\begin{tcolorbox}
    {\fontfamily{cmss}\selectfont
    \textbf{class} coupled\_cluster.cc.\textbf{CoupledCluster}

    \hspace{1em}(\emph{system}, \emph{mixer=<class'coupled\_cluster.mix.DIIS'>}, 
        \emph{verbose=False}, \emph{np=None})

    \vspace{1em}
    Abstract base class defining the basic structure of a coupled cluster ground state 
    solver class.
    
    \vspace{1em}
    \textbf{Parameters}

    \hspace{2em}\textbf{system}(\emph{QuantumSystem}) A system class from the 
        \emph{quantum\_systems} module.

    \hspace{2em}\textbf{mixer}(\emph{AlhaMixer, default AlphaMixer}) Mixer - 
        Subclass of \emph{AlphaMixer} class.

    \hspace{2em}\textbf{verbose}(\emph{bool, default False}) Will print results 
        for each iteration if \emph{True}.

    \vspace{1em}
    \textbf{Methods}

    \hspace{2em} \textbf{compute\_ground\_state}
        (\emph{t\_args=[]}, \emph{t\_kwargs=\{\}}, \emph{l\_argys=[]}, \emph{l\_kwargs=\{\}})
        \begin{adjustwidth}{4em}{}
        Computes ground state of system given as parameter. Allows for parameters relating 
        the the $\tau$- and $\lambda$ amplitudes, for use in inheriting classes.
        \end{adjustwidth}
    
    \hspace{2em} \textbf{compute\_particle\_density}()
        \begin{adjustwidth}{4em}{}
        Computes the one-body density of the system.
   
        \textbf{Returns:} Particle density 
  
        \textbf{Return type:} \emph{np.array}
        \end{adjustwidth}
    
    \hspace{2em} \textbf{compute\_reference\_energy}()
        \begin{adjustwidth}{4em}{}
        Computes reference energy

        \textbf{Returns:} Reference energy

        \textbf{Return type:} \emph{np.array}
        \end{adjustwidth}
    
    \hspace{2em} \textbf{get\_amplitudes}(\emph{get\_t\_0=False})
        \begin{adjustwidth}{4em}{}
        Getter for amplitudes.

        \textbf{Parameters:} \textbf{get\_t\_0} (\emph{bool, default False}) 
            Returns amplitude at $t=0$ if \lstinline{True}.

        \textbf{Returns:} Amplitudes

        \textbf{Return type:} \emph{AmplitudeContainer}

        \end{adjustwidth}

    }
\end{tcolorbox}