\begin{tcolorbox}
    {\fontfamily{cmss}\selectfont
    \textbf{class} coupled\_cluster.cc.\textbf{OACCD}
    (\emph{system}, \emph{**kwargs})

    \vspace{1em}
    Implementation of the orbital-adaptive coupled cluster method with double excitation,
    also called the nonorthogonal orbital-optimized coupled cluster model.
    Requires orthonormal basis functions. Based on work by Rolf H. 
    Myhre\cite{myhre2018demonstrating}.

    Inherits from the \textbf{CoupledCluster} abstract base class.

    \vspace{1em}
    \textbf{Parameters}

    \hspace{2em}\textbf{system}(\emph{QuantumSystem}) A system class from the 
        \emph{quantum\_systems} module.

    \vspace{1em} 
    \textbf{Methods}

    \hspace{2em} \textbf{compute\_ground\_state}
        (\emph{max\_iterations=$100$}, \emph{tol=$1e^{-4}$},

        \hspace{3em} \emph{termination\_tol=$1e^{-4}$}, \emph{tol\_factor=$0.1$},
        \emph{change\_system\_basis=False},
        
        \hspace{3.2em}\emph{**mixer\_kwargs})

        \begin{adjustwidth}{4em}{}
        Computes ground state of system.

        \textbf{Parameters:} 

            \hspace{1.5em}\textbf{max\_iterations} (\emph{int, default $100$}) 
            Maximum number of iterations.

            \hspace{1.5em}\textbf{tol} (\emph{float, default $1e^{-4}$})
            Tolerance of convergence.

            \hspace{1.5em}\textbf{termination\_tol} (\emph{float, default $1e^{-4}$})
            Give up if tolerance is below this.

            \hspace{1.5em}\textbf{tol\_factor} (\emph{float, default $0.1$})
            Decreases tolerance if non-convergent.

            \hspace{1.5em}\textbf{change\_system\_basis} (\emph{bool, default False})
            Changes basis after calculation.

        \end{adjustwidth}
    } 
\end{tcolorbox}