\begin{tcolorbox}
    {\fontfamily{cmss}\selectfont
    \textbf{class} coupled\_cluster.cc.\textbf{TDCCSD}
    (\emph{*args}, \emph{**kwargs})

    \vspace{1em}
    Sub-class of \textbf{TimeDependentCoupledCluster}
    Class for computing time-development of provided system, employing time-dependent 
    coupled cluster method with single- and double excitations. The orbitals are kept 
    static. This class inherits all methods from the parent class, but includes 
    a few extra.

    \vspace{1em}
    \textbf{Parameters}

    \hspace{2em}\textbf{cc}(\emph{CoupledCluster})
        Class instance defining the ground state solver.

    \hspace{2em}\textbf{system}(\emph{QuantumSystem}) 
        Class instance defining the system to be solved.

    \hspace{2em}\textbf{np}(\emph{module})
        Matrix/linear algebra library to be used, e.g. Numpy, Cupy
    
    \hspace{2em}\textbf{integrator}(\emph{Integrator})
        Integrator class instance, e.g. RK4, GaussIntegrator

    \vspace{1em}
    \textbf{Methods}

    \hspace{2em} \textbf{compute\_energy} ()
        \begin{adjustwidth}{4em}{}
        Computes energy at current time step.

        \textbf{Returns:} energy
        \textbf{Return type:} \emph{float}
        \end{adjustwidth}

    \hspace{2em} \textbf{compute\_time\_dependent\_overlap} ()
        \begin{adjustwidth}{4em}{}
        Computes overlap of current time-developed state with the ground state.

        \textbf{Returns:} Probability of grond state 
        \textbf{Return type:} \emph{np.array}
        \end{adjustwidth}

    } 
\end{tcolorbox}